---
title: "new_decision_tree_model"
author: "yana xu"
date: "2024-11-24"
output: html_document
---
## 1. Loading Libraries and Dataset
```{r setup}
# Load necessary libraries
library(mice)        # For multiple imputation
library(dplyr)       # For data manipulation
library(ggplot2)     # For data visualization
library(caret)       # For model training and evaluation
library(rpart)       # For building decision trees
library(rpart.plot)  # For plotting decision trees
library(pROC)        # For ROC curve analysis
library(ggcorrplot)  # For plotting correlation matrices

# Read the dataset
heart_data <- read.csv("~/Desktop/625_Final/dataset_cleaned/heart_disease_dataset.csv")

# Inspect the structure of the dataset
str(heart_data)

# View the first few rows
head(heart_data)

# Summary statistics
summary(heart_data)

# Check for missing values
colSums(is.na(heart_data))

```

## Data Preprocessing

```{r data process}
# Convert appropriate columns to factors
heart_data <- heart_data %>%
  mutate(
    sex = as.factor(sex),
    cp = as.factor(cp),
    fbs = as.factor(fbs),
    restecg = as.factor(restecg),
    exang = as.factor(exang),
    num = as.factor(ifelse(num > 0, 1, 0))  # 1 for disease, 0 for no disease
  )

# Replace invalid '0' values in 'chol' and 'trestbps' with NA
heart_data$chol[heart_data$chol == 0] <- NA
heart_data$trestbps[heart_data$trestbps == 0] <- NA

# Remove unnecessary columns
heart_data <- heart_data %>%
  select(-slope, -ca, -thal)

# Check missing values after preprocessing
colSums(is.na(heart_data))

```

## 4. Exploratory Data Analysis (EDA)

### 4.1 Age Distribution
```{r}
# Histogram of age
age_plot <- ggplot(heart_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = 'blue', color = 'black') +
  theme_minimal() +
  labs(title = 'Age Distribution', x = 'Age', y = 'Frequency')

# Save the plot
ggsave(filename = "~/Desktop/625_Final/Decision Tree Model/result/age_distribution.png", plot = age_plot, width = 8, height = 6)
```

### 4.2 Distribution of Heart Disease
```{r}
# Bar plot of target variable
heart_disease_plot <- ggplot(heart_data, aes(x = num)) +
  geom_bar(fill = 'orange') +
  theme_minimal() +
  labs(title = 'Distribution of Heart Disease', x = 'Heart Disease (0 = No, 1 = Yes)', y = 'Count')

# Save the plot
ggsave(filename = "~/Desktop/625_Final/Decision Tree Model/result/heart_disease_distribution.png", plot = heart_disease_plot, width = 8, height = 6)

```

### 4.3 Correlation Matrix
```{r}
# Correlation matrix for numeric variables
numeric_vars <- heart_data %>% select_if(is.numeric)
corr_matrix <- cor(numeric_vars, use = 'complete.obs')

# Plot the correlation matrix
corr_plot <- ggcorrplot(corr_matrix, lab = TRUE, title = "Correlation Matrix")
# Save the plot
ggsave(filename = "~/Desktop/625_Final/Decision Tree Model/result/correlation_matrix.png", plot = corr_plot, width = 8, height = 6)
```

## 5. Handle Missing Data with Multiple Imputation

```{r}
# Set seed for reproducibility
set.seed(123)

# Impute missing values using mice
imputed_data <- mice(heart_data, m = 5, method = 'pmm', seed = 123)

# Check the imputation methods used
imputed_data$method

# Density plots before and after imputation for 'chol'
densityplot(imputed_data, ~chol)

# Extract the first imputed dataset
complete_data <- complete(imputed_data, 1)

# Verify that there are no missing values
colSums(is.na(complete_data))

```

## 6. Split the Data into Training and Testing Sets

```{r}
# Set seed for reproducibility
set.seed(123)

# Create training (80%) and testing (20%) datasets
train_index <- createDataPartition(complete_data$num, p = 0.8, list = FALSE)
train_data <- complete_data[train_index, ]
test_data  <- complete_data[-train_index, ]

# Verify the proportion of target variable in both sets
prop.table(table(train_data$num))
prop.table(table(test_data$num))
```

## 7. Model Training with Cross-Validation and Hyperparameter Tuning

```{r}
# Define cross-validation method
train_control <- trainControl(method = "cv", number = 10)

# Define a grid of complexity parameters for tuning
cp_grid <- expand.grid(cp = seq(0.001, 0.05, by = 0.001))

# Train the decision tree model
set.seed(123)
model <- train(
  num ~ .,
  data = train_data,
  method = "rpart",
  trControl = train_control,
  tuneGrid = cp_grid,
  metric = "Accuracy"
)

# View the best tuning parameter
print(model$bestTune)

# Plot model performance vs. complexity parameter
model_performance_plot <- plot(model)
# Save the plot
png(filename = "~/Desktop/625_Final/Decision Tree Model/result/model_performance_plot.png", width = 800, height = 600)
plot(model)
dev.off()
```

## 8. Visualize the Final Decision Tree

```{r}
# Plot the final decision tree
png(filename = "~/Desktop/625_Final/Decision Tree Model/result/decision_tree.png", width = 800, height = 600)
rpart.plot(
  model$finalModel,
  type = 2,
  extra = 104,
  fallen.leaves = TRUE,
  main = "Decision Tree for Heart Disease Prediction"
)
dev.off()
```

## 9. Evaluate Model Performance on Test Data
```{r}
# Make predictions on the test set
predictions <- predict(model, test_data)

# Confusion matrix
conf_mat <- confusionMatrix(predictions, test_data$num)
print(conf_mat)
```

### Evaluation Metrics
```{r}
# Extract evaluation metrics
accuracy <- conf_mat$overall['Accuracy']
precision <- conf_mat$byClass['Pos Pred Value']
recall <- conf_mat$byClass['Sensitivity']
f1_score <- 2 * (precision * recall) / (precision + recall)

# Print evaluation metrics
cat("Accuracy: ", round(accuracy, 4), "\n")
cat("Precision: ", round(precision, 4), "\n")
cat("Recall: ", round(recall, 4), "\n")
cat("F1 Score: ", round(f1_score, 4), "\n")
```

## 10. ROC Curve and AUC
```{r}
# Convert factors to numeric for ROC analysis
roc_predictions <- as.numeric(as.character(predictions))
roc_test_labels <- as.numeric(as.character(test_data$num))

# Compute ROC curve
roc_obj <- roc(roc_test_labels, roc_predictions)

# Plot ROC curve
png(filename = "~/Desktop/625_Final/Decision Tree Model/result/roc_curve.png", width = 800, height = 600)
plot(roc_obj, col = "blue", main = "ROC Curve for Decision Tree Model")
abline(a = 0, b = 1, lty = 2, col = "red")
dev.off()

# Calculate AUC
auc_value <- auc(roc_obj)
cat("AUC: ", round(auc_value, 4), "\n")
```

## 11. Variable Importance
```{r}
# Assess variable importance
importance <- varImp(model, scale = FALSE)
print(importance)

# Plot variable importance
importance_plot <- plot(importance, main = "Variable Importance")

# Save the plot
png(filename = "~/Desktop/625_Final/Decision Tree Model/result/variable_importance.png", width = 800, height = 600)
plot(importance, main = "Variable Importance")
dev.off()
```

## 12. Save the Model
```{r}
# Save the model to an RDS file
saveRDS(model, file = "~/Desktop/625_Final/Decision Tree Model/result/decision_tree_model.rds")
```
